---
title: "Graham et al 2021 LDL-C PRS Performance in Samoans"
author: "Jenna C. Carlson"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output: 
 html_document:
    toc: true
    number_sections: true
    code_folding: show
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

Goal: assess the performance of the multi-ancestry polygenic risk score
for LDL-C constructed in Graham et al 2021
(<https://www.nature.com/articles/s41586-021-04064-3>) in the Samoan
discovery cohort.

PRS weights from: <https://www.pgscatalog.org/score/PGS000889/>

This PRS was developed using GWAS from ALL ancestries in the analyses
(1,088,526 individuals of the following ancestries: African unspecified,
East Asian, European, Hispanic or Latin American, South Asian).

It was developed using pruning and thresholding of GWAS results (r2=0.1,
distance=500kb, p=5e-4) and contains 9,009 variants in build hg19,
weighted by effect estimates (betas).

# Load Libraries

```{r load_libraries,message=FALSE}
library(tidyverse)
library(SeqArray)
library(SeqVarTools)
library(TopmedPipeline)
library(Biobase) # annotated data frame
library(rsq)
library(RNOmni)
library(readxl)

```

# Input directory and files

```{r}
# Print the working directory
getwd()
```

Symbolic links

```{bash, eval=FALSE}
# Update paths to your local GDS, phenotype, and population structure files
#ln -s /path/to/freeze8-imputed/gds gds
ln -s /path/to/freeze9b/discovery/14_make_gds gds
ln -s /path/to/population_structure/data/samoa-hg38_pcair-round2_pcair.RData ./samoa-hg38_pcair-round2_pcair.RData
ln -s /path/to/Cholesterol2-model1/Cholesterol2-phenotype.RData
ln -s /path/to/phenotypes/phenotypes-analysis-discovery.RData
```

# Create pgs in master RMD
# Create scoring file is a script 
# Assign scores in a script

#Data analysis
## What percent of PRS variants have MAF = 0 in Samoans?

```{r}
maf.counts <- data.frame(chr = 1:22,
                         n.snp = rep(NA, 22),
                         n.maf0 = rep(0, 22))
for(i in 1:22){
  load(paste0("~/PRS/Total_chol/scoring-files/final.scoring.file.chr", i, ".RData"), verbose=FALSE)
  maf.counts$n.snp[i] <- nrow(final.scoring.file)
  maf.counts$n.maf0[i] <- sum(final.scoring.file$maf == 0, na.rm = T) # 2 vars on chr 19 have missing MAF info
}

sum(maf.counts$n.snp)
sum(maf.counts$n.maf0)
sum(maf.counts$n.maf0)/sum(maf.counts$n.snp)
```

# PRS evaluation in Samoans

## Merge in all chromosomes together

```{r}
load("~/PRS/Total_chol/prs-samoan-scores/samoa.prs.chr1.RData")
names(samoa.scores)[2] <- "score.chr1"

samoa.scores.all <- samoa.scores 


for(chr.i in c(2:22)){
  load(paste0("~/PRS/Total_chol/prs-samoan-scores/samoa.prs.chr", chr.i, ".RData"))
  names(samoa.scores)[2] <- paste0("score.chr", chr.i)
  samoa.scores.all <- left_join(samoa.scores.all, samoa.scores, by = "ID")
}

dim(samoa.scores.all)

samoa.scores.all$prs <- rowSums(samoa.scores.all[,-1])
summary(samoa.scores.all$prs)
sd(samoa.scores.all$prs)
```

## Distribution of PRS in Samoans

```{r}
# Update path to your local phenotype data file
load("path/to/phenotypes-analysis-discovery.RData")

# generate residuals
chol.obs <- complete.cases(phenotypes$age.chol, phenotypes$sex, phenotypes$chol)
m.chol <- lm(chol ~ age.chol + age2.chol + sex + sex_age.chol + sex_age2.chol + PC1 + PC2 + PC3, data = phenotypes[chol.obs,], na.action = na.exclude) 
phenotypes[chol.obs, "chol.res"] <- RankNorm(residuals(m.chol))
hist(phenotypes$chol.res)

#ggplot(data=samoa.scores.all, aes(x=prs)) + 
#  geom_density() + 
#  theme_bw()
```

## Association with LDL

Merge with phenotype data

```{bash, eval=FALSE}
# Update paths to your local phenotype and population structure files
#ln -s /path/to/LDL_C2-model1/LDL_C2-phenotype.RData

ln -s /path/to/Cholesterol2-model1/Cholesterol2-phenotype.RData

ln -s /path/to/population_structure/data/samoa-hg38-pcair-round2_pcair.RData



```

```{r}
load("Cholesterol2-phenotype.RData", verbose=TRUE)

head(pData(annot))

pca <- get(load("./samoa-hg38_pcair-round2_pcair.RData"))
pcs <- as.data.frame(pca$vectors)
pcs$sample.id <- row.names(pcs)

tmp.df <- left_join(samoa.scores.all, pData(annot), by = c("ID" = "sample.id"))
samoa.prs.df <- left_join(tmp.df, pcs[,c("sample.id", "V1", "V2", "V3")], by = c("ID" = "sample.id"))

samoa.prs.df
```

### Unadjusted model

```{r}
ggplot(data = samoa.prs.df, aes(y=Cholesterol2, x=prs)) + 
  geom_point(alpha=0.3, color="#004D85") + 
  geom_smooth(color="gray20", se=FALSE) + 
  theme_classic() +
  xlab("Polygenic Risk Score") + 
  ylab("Total Cholesterol (mg/dL)") +
  scale_x_continuous(breaks=seq(43,50)) +
  theme(text=element_text (size=32))

ggsave("PRS-Cholesterol2_scatterplot.png", width=16, height=10, units="in")


m1 <- lm(Cholesterol2 ~ prs, data = samoa.prs.df)
summary(m1)

```

didn't end up using this. ignore stuff with norm

```{r}
samoa.prs.df$PRS_norm <- RankNorm(samoa.prs.df$prs)
m1.n <- lm(Cholesterol2 ~ PRS_norm, data = samoa.prs.df)
summary(m1.n)
```

### Adjust for age, sex, and PCAs

```{r}
m2 <- lm(Cholesterol2 ~ prs + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = samoa.prs.df)
summary(m2)
confint(m2)

# p-value for prs
beta <- summary(m2)$coefficients["prs", "Estimate"]
beta
confint(m2)
tstat <- summary(m2)$coefficients["prs", "t value"]
2*pt(abs(tstat), df=2849, lower.tail=FALSE)

#m2.n <- lm(Cholesterol2 ~ PRS_norm + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = samoa.prs.df)
#summary(m2.n)
```

### R2 - PRS evaluation metric

```{r}
rsq.partial(objF = m2)

rsq.partial(objF = m2)$partial.rsq[1]

#rsq.partial(objF = m2.n)

#rsq.partial(objF = m2.n)$partial.rsq[1]
```

**The Partial R2 for the PRS in Samoans in 10.2%.**

Evaluation of PRS in Graham et al 2021:

"For each individual in the testing cohorts, PRS were calculated as the
sum of the dosages multiplied by the given weight at each variant."

"Model fit was assessed using the adjusted R2 of a linear model for
LDL-C value at initial assessment adjusted for cholesterol medication
(divided by 0.7 to estimate pre-medication levels) with sex, batch, age
at initial assessment and PCs 1--4 as covariates (Supplementary Tables
21--23)."

"Supplementary Table 23: Prediction of LDL-C in MGI individuals based on
varying numbers of PCs included in the model The polygenic score was
normalized within each ancestry group separately"

"Adjusted R2 values were reported for each cohort and ancestry group,
with 95% confidence intervals for the adjusted R2 values calculated
using bootstrapping."

"The PRS developed from the multi-ancestry meta-analysis consistently
showed the best or near-best performance in each group tested, with
improved or comparable predictions relative to ancestry-matched scores
(adjusted R2=0.10--0.16; Fig. 3, Supplementary Table 17, Extended Data
Fig. 7)."

#### Bootstrap CI for R2

```{r}
set.seed(06082022)
n.boot <- 1000
boot.r2 <- rep(NA, times=n.boot)

for(i in 1:n.boot){
  boot.df <- samoa.prs.df[sample(1:nrow(samoa.prs.df), nrow(samoa.prs.df), replace=TRUE),]
  boot.m2 <- lm(Cholesterol2 ~ prs + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = boot.df)
  boot.r2[i] <- rsq.partial(objF = boot.m2)$partial.rsq[1]
}

hist(boot.r2)
quantile(boot.r2, c(0.025, 0.975)) # 95% bootstrap CI for R2



```



# Session Information

```{r}
sessionInfo()
```
