---
title: "Graham et al 2021 LDL-C PRS Performance in Samoans"
author: "Toni-Ann Yapp and Jenna C. Carlson"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output: 
 html_document:
    toc: true
    number_sections: true
    code_folding: show
editor_options: 
  chunk_output_type: inline
---


```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```


Goal: assess the performance of the multi-ancestry polygenic risk score for LDL-C constructed in Graham et al 2021 (https://www.nature.com/articles/s41586-021-04064-3) in the Samoan discovery cohort.

PRS weights from: https://www.pgscatalog.org/score/PGS000889/

This PRS was developed using GWAS from ALL ancestries in the analyses (1,088,526 individuals of the following ancestries: African unspecified, East Asian, European, Hispanic or Latin American, South Asian).

It was developed using pruning and thresholding of GWAS results (r2=0.1, distance=500kb, p=5e-4) and contains 9,009 variants in build hg19, weighted by effect estimates (betas).


# Load Libraries

```{r load_libraries,message=FALSE}
library(tidyverse)
library(SeqArray)
library(SeqVarTools)
library(TopmedPipeline)
library(Biobase) # annotated data frame
library(rsq)
library(RNOmni)
library(readxl)

```

# Input directory and files 

```{r}
# Print the working directory
getwd()
```
Symbolic links

```{bash, eval=FALSE}
#ln -s /home/jzz10/gwas/freeze8-imputed/gds gds
ln -s /home/mok36/imputation_code/freeze9b/discovery/14_make_gds gds
ln -s /home/jnc35/gwas/imputed/imputed_freeze9/discovery/population_structure/data/samoa-hg38_pcair-round2_pcair.RData ./samoa-hg38_pcair-round2_pcair.RData
```

Get pgs from online. get pgs for each trait 
```{bash, eval=FALSE}
wget https://ftp.ebi.ac.uk/pub/databases/spot/pgs/scores/PGS002781/ScoringFiles/PGS002781.txt.gz
```


```{bash}
zcat PGS002781.txt.gz | head -50
```


```{r}
hdl_prs <- read_table(file = "PGS002781.txt.gz", comment = "#")

# any CHR X? (answer: no)
table(hdl_prs$chr_name, useNA="always")


```

# Liftover PRS variants (hg19 to hg38)

Make BED format file

http://genome.ucsc.edu/FAQ/FAQformat.html#format1

```{r}
hdl_bed <- hdl_prs %>% mutate(chrom = paste0("chr",chr_name),
                      chromStart = chr_position - 1,
                      chromEnd = chr_position)
head(hdl_bed)

hdl_bed %>% 
  dplyr::select(chrom, chromStart, chromEnd) %>% 
  write_delim(., file = "HDL_PRS_for_liftover.txt", delim = " ", col_names = FALSE)
```

http://genome.ucsc.edu/cgi-bin/hgLiftOver
Original: Human hg19 
New: Human hg38


```{r}
deleted= read.table("genome.ucsc.edu_trash_hglft_genome_3e84b_9d2400.err.txt", comment = "#") 

bed_hdl_hg19_new = hdl_bed %>% anti_join(., deleted, by = c("chrom" = "V1", "chromStart" = "V2", "chromEnd"="V3"))
hdl_bed_inner_join_deleted = hdl_bed %>% inner_join(., deleted, by = c("chrom" = "V1", "chromStart" = "V2", "chromEnd"="V3"))


```

#Input deleted SNP

```{r}
hdl_prs_hg38 <- read_table("hglft_genome_3400a_9cd990.bed", col_names=c("chrom.hg38", "chromStart.hg38", "chromEnd.hg38", "x", "y"))

hdl_scores <- bed_hdl_hg19_new %>% add_column(hdl_prs_hg38)

```
# add back in failed SNP conversion

# find which row it should be
idx <- which(bed$chrom == "chr19" & bed$chromStart == 54763072 & bed$chromEnd == 54763073)

# insert new row by shifting rows down by 1
prs_hg38[(idx + 1):(nrow(prs_hg38) + 1), ] <- prs_hg38[idx:nrow(prs_hg38), ]
prs_hg38[idx, ] <- list("chr19", 54259223, 54259224)

# double check
prs_hg38[idx:(idx+3), ]

table(duplicated(prs_hg38[, c("chrom", "chromStart")]))


# find which row it should be
idx <- which(bed$chrom == "chr19" & bed$chromStart == 54756141 & bed$chromEnd == 54756142)

# insert new row by shifting rows down by 1
prs_hg38[(idx + 1):(nrow(prs_hg38) + 1), ] <- prs_hg38[idx:nrow(prs_hg38), ]
prs_hg38[idx, ] <- list("chr19", 4252278, 4252279)

# double check
prs_hg38[idx:(idx+3), ]



Merge back in with scores

```{r}
#hdl_prs_hg38 <-hdl_prs_hg38 %>% rename_with(~ paste0(.x, ".hg38"))
#hdl_scores <- bed_hdl_hg19_new %>% add_column(hdl_prs_hg38)
```

# Extract variants from Freeze9 GDS files

Create a data frame with genotypes for all 9,009 variants for the score

```{r}
table(hdl_scores$chr_name) # all chromosomes
```

## Example: Chromosome 1 

```{r, eval=FALSE}
# chr 1
tmp.scores <- hdl_scores %>% filter(chr_name == 1)

# extract just the PRS variants from this chr
#gds <- seqOpen("gds/samoa-chr1.dose_10a") #freeze 8
gds <- seqOpen("gds/discovery-9b-samoa-chr1.dose.gds")  # freeze 9 
pos <- seqGetData(gds, "position")
seqSetFilter(gds, variant.sel = (pos %in% tmp.scores$chromEnd.hg38))
pos <- seqGetData(gds, "position")

# extract alleles from PRS variants (format: "ref,alt")
alleles <- seqGetData(gds, "allele")
ref <- sapply(1:length(alleles), function(i) strsplit(alleles[i], ",")[[1]][1])
alt <- sapply(1:length(alleles), function(i) strsplit(alleles[i], ",")[[1]][2])

maf <- seqGetData(gds, "annotation/info/MAF")

samoa.vars <- tibble(var.id = seqGetData(gds, "variant.id"), # matches column name in dos (below)
                     chr = rep(1, length(pos)), 
                     pos = pos,
                     ref = ref, 
                     alt = alt,
                     maf = maf)

samoa.vars

# get genotypes (imputed dosages)
dos <- imputedDosage(gds, dosage.field="DS", use.names=TRUE) %>% as.data.frame() %>% rownames_to_column(., "ID") 
dos[1:5,1:5]

seqClose(gds)

```


### Investigate the variants from the PRS that are not in the GDS file

```{r}
tmp.scores$inGDS <- FALSE
tmp.scores$inGDS[tmp.scores$chromEnd.hg38 %in% pos] <- TRUE

table(tmp.scores$inGDS)


miss.pos <- tmp.scores %>% filter(inGDS == FALSE) %>% mutate(SNP = paste0(chr_name, ":", chromEnd.hg38))
miss.pos
```


Look for variants in imputation working files

```{bash}
#grep -m 1 '1:56105434' /home/jzz10/gwas/freeze8-imputed/freeze8-samoa-hg38-chr1-imputed.info
grep -m 1 '1:63199348' /home/mok36/imputation_code/freeze9b/discovery/9_Minimac4_imputation/discovery-9b-hg38-chr1-imputed.info
```

Check if all these variants had MAF = 0 and that's why they're excluded (not run with freeze 9 as maf=0 variants are now included in the freeze 9 gds files)

```{r, eval=FALSE}
fileConn <- file("find-missing-PRS-variants-chr1.sh")

writeLines(c("#!/bin/sh",
             paste0("grep -m 1 '\\<", miss.pos$SNP, "\\>' /home/jzz10/gwas/freeze8-imputed/freeze8-samoa-hg38-chr1-imputed.info")), 
           # m -1 means spit out the first result and then stop
           # \<stuff\>' means match exactly that (not stuff with extra digits)
           fileConn)

close(fileConn)
```


```{bash, eval=FALSE}
/bin/sh find-missing-PRS-variants-chr1.sh > info-chr1.info
```

```{r, eval=FALSE}
info.chr1 <- read_table("info-chr1.info", 
                        col_names = c("SNP", "REF0", "ALT1", "ALT_Frq", "MAF", "AvgCall", "Rsq", "Genotyped"),
                        col_types = "cccddddc-----",
                        na="-")

table(info.chr1$MAF) # Check all MAF is 0

nrow(info.chr1) == nrow(miss.pos) # check that it found all variants
nrow(miss.pos) - nrow(info.chr1) # number of SNPs missing



# identify which are still missing
info.chr1$POS <- sapply(info.chr1$SNP, function(x) str_split(x, ":")[[1]][2])

tmp.scores$MAF0 <- FALSE
tmp.scores$MAF0[(tmp.scores$chromEnd.hg38 %in% info.chr1$POS)] <- TRUE
table(tmp.scores$MAF0)

still.missing <- tmp.scores %>% filter(!MAF0 & !inGDS) 
still.missing <- still.missing %>% mutate(chr_sites = paste0("chr", chr_name)) # sites files use chr prefix

still.missing

write_delim(still.missing %>% dplyr::select(chr_sites, chromEnd.hg38), 
            file = "find-missing-PRS-variants-from-sites-chr1.region", 
            col_names = FALSE,
            delim="\t")
```

look for missing variants in the sites files  

```{r}
still.missing <- miss.pos %>% mutate(chr_sites = paste0("chr", chr_name)) # sites files use chr prefix

write_delim(still.missing %>% dplyr::select(chr_sites, chromEnd.hg38), 
            file = "find-missing-PRS-variants-from-sites-chr1-f9.region", 
            col_names = FALSE,
            delim="\t")
```

```{bash, eval=FALSE}
bcftools view -R find-missing-PRS-variants-from-sites-chr1-f9.region /home/shared_data/samoa/wgs/freeze.9b/sites/freeze.9b.chr1.pass_and_fail.sites.bcf -Ov -o freeze.9b.sites.missing.vcf
bcftools query -f'%CHROM  %POS  %REF  %ALT %QUAL %FILTER %INFO/AC\n' freeze.9b.sites.missing.vcf -o freeze.9b.sites.missing.txt
```

```{r}
sites.file.miss <- read_table("freeze.9b.sites.missing.txt", 
                              col_names = c("CHROM", "POS", "REF", "ALT", "QUAL", "FILTER", "AC"))

table(sites.file.miss$FILTER) # most variables failed filter

tmp.scores$FAILED_FILTER <- FALSE
tmp.scores$FAILED_FILTER[(tmp.scores$chromEnd.hg38 %in% sites.file.miss$POS)] <- TRUE
table(tmp.scores$FAILED_FILTER)

#table(colSums(tmp.scores[, c("inGDS", "MAF0", "FAILED_FILTER")])) # any still missing (column sum of 0)? no (freeze 8)
table(colSums(tmp.scores[, c("inGDS", "FAILED_FILTER")])) # any still missing (column sum of 0)? no (freeze 9)
```

Summarize investigation of variants on chr 1

```{r}
nrow(tmp.scores) # number of variants in PRS on chr1

table(tmp.scores$inGDS) # number of variants available in Samoan GDS (test data)
table(tmp.scores$inGDS) %>% prop.table() 

maf0 <- samoa.vars %>% filter(maf == 0)
tmp.scores$MAF0 <- FALSE
tmp.scores$MAF0[(tmp.scores$chromEnd.hg38 %in% maf0$pos)] <- TRUE

table(tmp.scores$MAF0) # number of variants that are monomorphic in Samoans (test data)
table(tmp.scores$MAF0) %>% prop.table()

table(tmp.scores$FAILED_FILTER) # number of variants that failed QC in Samoans (test data)
table(tmp.scores$FAILED_FILTER) %>% prop.table()

```

**Freeze 9** From chr 1, there are 1,102 variants in++++++++++++++
the PRS; of those 1068 (97%) are present in the Samoan GDS files and 34 (3%) failed QC. There are 822 polymorphic variants in the Samoan data, with 280 (25%) monomorphic.

**Freeze 8** From chr 1, there are 1,102 variants in the PRS. Of those 691 (63%) are present and polymorphic in the Samoan data, 389 (35%) are monomorphic in the Samoan data, and 22 (2%) failed QC in the Samoan data. 

### Apply PRS to Samoan data

### Create Scoring File
```{r}
dim(tmp.scores) # PRS variants

dim(samoa.vars) # Samoan variants
samoa.vars

# check allele agreement
scoring.file <- inner_join(samoa.vars, tmp.scores[, c("effect_allele", "other_allele", "effect_weight", "chromEnd.hg38")], by = c("pos" = "chromEnd.hg38"))

scoring.file

scoring.file$alleles.match <- FALSE
scoring.file$alleles.match[scoring.file$alt == scoring.file$effect_allele & scoring.file$ref == scoring.file$other_allele] <- TRUE
table(scoring.file$alleles.match)

# create column to indicate whether to flip dosage (2 -> 0, etc) when generating PRS
scoring.file$alleles.flip <- FALSE
scoring.file$alleles.flip[scoring.file$alt == scoring.file$other_allele & scoring.file$ref == scoring.file$effect_allele] <- TRUE
table(scoring.file$alleles.flip)

# double check no variant has both
xtabs(~ alleles.flip + alleles.match, data = scoring.file)

# non-matching alleles
scoring.file %>% filter(!alleles.flip &!alleles.match)

# possible strand flip (A -> T, C -> G, G -> C, T -> A)
strand_flip <- function(allele){
  if (allele == "A") {
    flip <- "T"
  } else if (allele == "C") {
    flip <- "G"
  } else if (allele == "G") {
    flip <- "C"
  } else if (allele == "T") {
    flip <- "A"
  }
  return(flip)
}

View(scoring.file)
scoring.file$flip.strand <- FALSE 

for(i in 1:nrow(scoring.file)){
  if(!isTRUE(scoring.file$alleles.match[i]) & !isTRUE(scoring.file$alleles.flip[i])) {
    if (isTRUE(scoring.file$alt[i] == strand_flip(scoring.file$effect_allele[i])) & 
        isTRUE(scoring.file$ref[i] == strand_flip(scoring.file$other_allele[i]))) {
      scoring.file$flip.strand[i] <- TRUE
    }
  }
}

table(scoring.file$flip.strand)

scoring.file$flip.strand.swap.alleles <- FALSE
for(i in 1:nrow(scoring.file)){
  if(!scoring.file$alleles.match[i] & !scoring.file$alleles.flip[i] & !scoring.file$flip.strand[i]){
    if (scoring.file$alt[i] == strand_flip(scoring.file$other_allele[i]) & scoring.file$ref[i] == strand_flip(scoring.file$effect_allele[i])){
      scoring.file$flip.strand.swap.alleles[i] <- TRUE
    }
  }
}

table(scoring.file$flip.strand.swap.alleles)

# flag to exclude a variant for mismatched alleles
scoring.file$drop <- FALSE
scoring.file$drop[!scoring.file$alleles.match & 
                  !scoring.file$alleles.flip & 
                  !scoring.file$flip.strand & 
                  !scoring.file$flip.strand.swap.alleles] <- TRUE
table(scoring.file$drop)

final.scoring.file <- scoring.file %>% filter(!drop)
```

### Assign Individual Scores
```{r, eval=FALSE}
# subset dos to match the final scoring file
final.dos <- dos %>% dplyr::select(ID, as.character(final.scoring.file$var.id))

# check that they are lined up in the same variant order
table(names(final.dos)[2:ncol(final.dos)] == as.character(final.scoring.file$var.id))

chr1.scores <- tibble(ID = final.dos$ID,
                      score = 0)

head(chr1.scores)

# Flip dosages for variants that have allele flips
cols.to.flip <- final.scoring.file %>% filter(alleles.flip | flip.strand.swap.alleles) %>% dplyr::pull(var.id) %>% as.character
final.dos[,cols.to.flip] <- (2 - final.dos[,cols.to.flip])

# For each person, extract the vector of dosages and multiple by the vector of effect weights
chr1.weights <- final.scoring.file %>% dplyr::pull(effect_weight)
for(person.i in 1:nrow(chr1.scores)){
  chr1.scores$score[person.i] <- sum(chr1.weights * final.dos[person.i, -1 ]) 
}

summary(chr1.scores$score)

```

# Repeat process for all chromosomes

## Extract variants from Freeze9 GDS files and create scoring files

Updated to (1) remove non SGIDs from the gds file before extracting dosages, and (2) imputed to the mean dosage for any variants with missing genotypes (those that were dropped from the WGS data due to QC [filter==SVM, etc.] but that were genotyped and had a handful of missing genotypes)

```{r, eval=FALSE}
for(chr.i in c(1:22)){
  
  cat("\n\n")
  cat(paste0("extracting variants for chromosome ", chr.i))
  cat("\n")

  # extract scoring info for that chr
  tmp.scores <- scores %>% filter(chr_name == chr.i)

  # extract just the PRS variants from this chr
  #gds <- seqOpen(paste0("gds/samoa-chr", chr.i, ".dose_10a.gds"))
  gds <- seqOpen( paste0("gds/discovery-9b-samoa-chr", chr.i, ".dose.gds"))  # freeze 9 

  # drop non-SGIDs
  sample.id <- seqGetData(gds, "sample.id")
  keep.id <- sample.id[grep("^SG", sample.id)]
  seqSetFilter(gds, sample.id=keep.id)
  
  # extract variants by position in scoring file
  pos <- seqGetData(gds, "position")
  seqSetFilter(gds, variant.sel = (pos %in% tmp.scores$chromEnd.hg38))
  pos <- seqGetData(gds, "position")
  
  # extract alleles from PRS variants (format: "ref,alt")
  alleles <- seqGetData(gds, "allele")
  ref <- sapply(1:length(alleles), function(i) strsplit(alleles[i], ",")[[1]][1])
  alt <- sapply(1:length(alleles), function(i) strsplit(alleles[i], ",")[[1]][2])
  
  maf <- seqGetData(gds, "annotation/info/MAF")

  samoa.vars <- tibble(var.id = seqGetData(gds, "variant.id"), # matches column name in dos (below)
                     chr = rep(chr.i, length(pos)), 
                     pos = pos,
                     ref = ref, 
                     alt = alt,
                     maf = maf)

  
  # get genotypes (imputed dosages)
  dos <- imputedDosage(gds, dosage.field="DS", use.names=TRUE) %>% as.data.frame() %>% rownames_to_column(., "ID") 
  seqClose(gds)

  # mean impute missingness
  cols.to.impute <- which(colSums(is.na(dos))>0) %>% unname()
  for(col.i in cols.to.impute) {
    dos[which(is.na(dos[, col.i])), col.i] <- mean(dos[,col.i], na.rm = TRUE)
  }
  
  
  # count how many variants found
  tmp.scores$inGDS <- FALSE
  tmp.scores$inGDS[tmp.scores$chromEnd.hg38 %in% pos] <- TRUE
  
  cat(paste0("variants located on chr ", chr.i))
  print(table(tmp.scores$inGDS))
  cat("\n\n")

  ### Harmonization between Samoan variants and PRS variants

  # check allele agreement
  scoring.file <- inner_join(samoa.vars, tmp.scores[, c("effect_allele", "other_allele", "effect_weight", "chromEnd.hg38")], by = c("pos" = "chromEnd.hg38"))
  
  scoring.file$alleles.match <- FALSE
  scoring.file$alleles.match[scoring.file$alt == scoring.file$effect_allele & scoring.file$ref == scoring.file$other_allele] <- TRUE
  table(scoring.file$alleles.match)
  
  # create column to indicate whether to flip dosage (2 -> 0, etc) when generating PRS
  scoring.file$alleles.flip <- FALSE
  scoring.file$alleles.flip[scoring.file$alt == scoring.file$other_allele & scoring.file$ref == scoring.file$effect_allele] <- TRUE
  
  cat(paste0("allele flips on chr ", chr.i))
  print(table(scoring.file$alleles.flip))
  cat("\n\n")

  # create column to indicate strand flips
  scoring.file$flip.strand <- FALSE
  for(i in 1:nrow(scoring.file)){
    if(!scoring.file$alleles.match[i] & !scoring.file$alleles.flip[i]){
      if (scoring.file$alt[i] == strand_flip(scoring.file$effect_allele[i]) & scoring.file$ref[i] == strand_flip(scoring.file$other_allele[i])){
        scoring.file$flip.strand[i] <- TRUE
      }
    }
  }
  
  cat(paste0("strand flips on chr ", chr.i))
  print(table(scoring.file$flip.strand))
  cat("\n\n")
  
  # create column for a flipped strand with allele swap
  scoring.file$flip.strand.swap.alleles <- FALSE
  for(i in 1:nrow(scoring.file)){
    if(!scoring.file$alleles.match[i] & !scoring.file$alleles.flip[i] & !scoring.file$flip.strand[i]){
      if (scoring.file$alt[i] == strand_flip(scoring.file$other_allele[i]) & scoring.file$ref[i] == strand_flip(scoring.file$effect_allele[i])){
        scoring.file$flip.strand.swap.alleles[i] <- TRUE
      }
    }
  }
  
  cat(paste0("strand flips with allele swap on chr ", chr.i))
  print(  table(scoring.file$flip.strand.swap.alleles))
  cat("\n\n")
  
  # flag to exclude a variant for mismatched alleles
  scoring.file$drop <- FALSE
  scoring.file$drop[!scoring.file$alleles.match & 
                    !scoring.file$alleles.flip & 
                    !scoring.file$flip.strand & 
                    !scoring.file$flip.strand.swap.alleles] <- TRUE
  
  cat(paste0("variants with unreconcilable alleles to be dropped on chr ", chr.i))
  print(table(scoring.file$drop))
  cat("\n\n")
  
  
  # create final scoring file for chr
  final.scoring.file <- scoring.file %>% filter(!drop)

  # subset dos to match the final scoring file
  final.dos <- dos %>% dplyr::select(ID, as.character(final.scoring.file$var.id))
  
  # check that they are lined up in the same variant order
  cat(paste0("check variants align between scoring and dos files on chr ", chr.i))
  print(table(names(final.dos)[2:ncol(final.dos)] == as.character(final.scoring.file$var.id)))
  cat("\n\n") 
  
  # Flip dosages for variants that have allele flips
  cols.to.flip <- final.scoring.file %>% filter(alleles.flip | flip.strand.swap.alleles) %>% dplyr::pull(var.id) %>% as.character
  final.dos[,cols.to.flip] <- (2 - final.dos[,cols.to.flip])

  save(final.scoring.file, file=paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"))
  save(final.dos, file=paste0("dosages/final.dos.chr", chr.i))

  rm(tmp.scores, samoa.vars, dos, scoring.file, final.scoring.file, final.dos)
}


```

## Assign Individual Scores
multiplym dosage and weights and add across chromosomes and add back into scoring file 

```{r}
for(chr.i in c(1:22)){
  cat("\n\n")
  cat(paste0("assigning scores for chromosome ", chr.i))
  cat("\n")
  
  load(paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"))
  load(paste0("dosages/final.dos.chr", chr.i))
  
  samoa.scores <- tibble(ID = final.dos$ID,
                      score = 0)
  
  # For each person, extract the vector of dosages and multiple by the vector of effect weights
  weights <- final.scoring.file %>% dplyr::pull(effect_weight)
  
  for(person.i in 1:nrow(samoa.scores)){
    samoa.scores$score[person.i] <- sum(weights * final.dos[person.i, -1 ]) 
  }
  
  cat(paste0("Summary of PRS on chr ", chr.i))
  cat("\n")
  print(summary(samoa.scores$score))
  cat("\n\n") 

  save(samoa.scores, file=paste0("prs-samoan-scores/samoa.prs.chr", chr.i, ".RData"))
  rm(samoa.scores)

}


```

### freeze 9 NA checks

#### check chr 13 (freeze 9 only)

assigning scores for chromosome 13
Summary of PRS on chr 13
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
 0.7965  0.9931  1.0408  1.0415  1.0886  1.2836    2504 
 
```{r, eval=FALSE}
chr.i <- 13

  load(paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"), verbose=TRUE)
  load(paste0("dosages/final.dos.chr", chr.i))
  
  samoa.scores <- tibble(ID = final.dos$ID,
                      score = 0)
  
  # For each person, extract the vector of dosages and multiple by the vector of effect weights
  weights <- final.scoring.file %>% dplyr::pull(effect_weight)
  
  for(person.i in 1:nrow(samoa.scores)){
    samoa.scores$score[person.i] <- sum(weights * final.dos[person.i, -1 ]) 
  }
  
  summary(samoa.scores$score)

samoa.scores %>% filter(is.na(score)) %>% head()

summary(weights)

# check for missing dosages
#apply(final.dos[, -1], 2, summary)
names(which(colSums(is.na(final.dos))>0))
summary(final.dos[, names(which(colSums(is.na(final.dos))>0))]) 

# print IDs with missing dosage
final.dos$ID[which(is.na(final.dos[, names(which(colSums(is.na(final.dos))>0))]))]


# do all missing dosages have HG IDs (1000G)? (no)
length(grep("HG", 
           final.dos$ID[which(is.na(final.dos[, names(which(colSums(is.na(final.dos))>0))]))]))

# exclude 1000G (HG ids) and then print IDs with missing dosage
no1000g <- final.dos[!str_detect(final.dos$ID, "HG"), ] 
no1000g$ID[which(is.na(no1000g[, names(which(colSums(is.na(no1000g))>0))]))]

# exclude 1000G (HG ids) and "NA" ids and then print IDs with missing dosage
no1000g2 <- no1000g[!str_detect(no1000g$ID, "NA"), ] 
no1000g2$ID[which(is.na(no1000g2[, names(which(colSums(is.na(no1000g2))>0))]))]


# variant name
final.scoring.file %>% filter(var.id == names(which(colSums(is.na(final.dos))>0)))
```

#### check chr 14 (freeze 9 only)

assigning scores for chromosome 14
Summary of PRS on chr 14
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
-0.2953  0.0338  0.0698  0.0690  0.1024  0.2511    2530 


```{r, eval=FALSE}
chr.i <- 14

  load(paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"), verbose=TRUE)
  load(paste0("dosages/final.dos.chr", chr.i))
  
  samoa.scores <- tibble(ID = final.dos$ID,
                      score = 0)
  
  # For each person, extract the vector of dosages and multiple by the vector of effect weights
  weights <- final.scoring.file %>% dplyr::pull(effect_weight)
  
  for(person.i in 1:nrow(samoa.scores)){
    samoa.scores$score[person.i] <- sum(weights * final.dos[person.i, -1 ]) 
  }
  
  summary(samoa.scores$score)

samoa.scores %>% filter(is.na(score)) %>% head()

summary(weights)

# check for missing dosages
#apply(final.dos[, -1], 2, summary)
names(which(colSums(is.na(final.dos))>0))
summary(final.dos[, names(which(colSums(is.na(final.dos))>0))]) 

# print IDs with missing dosage
final.dos$ID[which(is.na(final.dos[, names(which(colSums(is.na(final.dos))>0))]))]


# do all missing dosages have HG IDs (1000G)? (no)
length(grep("HG", 
           final.dos$ID[which(is.na(final.dos[, names(which(colSums(is.na(final.dos))>0))]))]))

# exclude 1000G (HG ids) and then print IDs with missing dosage
no1000g <- final.dos[!str_detect(final.dos$ID, "HG"), ] 
no1000g$ID[which(is.na(no1000g[, names(which(colSums(is.na(no1000g))>0))]))]

# exclude 1000G (HG ids) and "NA" ids and then print IDs with missing dosage
no1000g2 <- no1000g[!str_detect(no1000g$ID, "NA"), ] 
no1000g2$ID[which(is.na(no1000g2[, names(which(colSums(is.na(no1000g2))>0))]))]
no1000g2$ID[which(is.na(no1000g2[, names(which(colSums(is.na(no1000g2))>0))]))] %>% length()

# variant name
final.scoring.file %>% filter(var.id == names(which(colSums(is.na(final.dos))>0)))
```


```{bash, eval=FALSE}
grep 70685060 /home/mok36/imputation_code/freeze9b/discovery/9_Minimac4_imputation/discovery-9b-hg38-chr14-imputed.info

bcftools view -r chr14:70685060 /home/shared_data/samoa/wgs/freeze.9b/sites/freeze.9b.chr14.pass_and_fail.sites.bcf -H

grep 70685060 /home/mok36/imputation_code/freeze9b/discovery/3_Affy_discovery_samoa_plink/discovery-samoa-hg38-chr14.bim

plink --bfile /home/mok36/imputation_code/freeze9b/discovery/3_Affy_discovery_samoa_plink/discovery-samoa-hg38-chr14 --freq --snp rs11845258 --out ~/lipids/prs/variant-checking/chr14_70685060
```

Variant with missing genotypes is one that was *not* imputed and was added back from the Affy data. 

The genotyping rate is 0.991664 which translates to exactly 26 missing genotypes.

#### check chr 19 (freeze 9 only)

assigning scores for chromosome 19
Summary of PRS on chr 19
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
  36.04   39.20   39.51   39.45   39.75   41.56    2556 

```{r, eval=FALSE}
chr.i <- 19


  load(paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"), verbose=TRUE)
  load(paste0("dosages/final.dos.chr", chr.i))
  
  samoa.scores <- tibble(ID = final.dos$ID,
                      score = 0)
  
  # For each person, extract the vector of dosages and multiple by the vector of effect weights
  weights <- final.scoring.file %>% dplyr::pull(effect_weight)
  
  for(person.i in 1:nrow(samoa.scores)){
    samoa.scores$score[person.i] <- sum(weights * final.dos[person.i, -1 ]) 
  }
  
  summary(samoa.scores$score)

samoa.scores %>% filter(is.na(score)) %>% head()

summary(weights)

# check for missing dosages
#apply(final.dos[, -1], 2, summary)
names(which(colSums(is.na(final.dos))>0))

### var 1
variant <- names(which(colSums(is.na(final.dos))>0))[1]
summary(final.dos[, variant]) 

# print IDs with missing dosage
final.dos$ID[which(is.na(final.dos[, variant]))]
final.dos$ID[which(is.na(final.dos[, variant]))] %>% length()

# do all missing dosages have HG IDs (1000G)? (no)
length(grep("HG", 
           final.dos$ID[which(is.na(final.dos[, variant]))]))

# exclude 1000G (HG ids) and then print IDs with missing dosage
no1000g <- final.dos[!str_detect(final.dos$ID, "HG"), ] 
no1000g$ID[which(is.na(no1000g[, variant]))]

# exclude 1000G (HG ids) and "NA" ids and then print IDs with missing dosage
no1000g2 <- no1000g[!str_detect(no1000g$ID, "NA"), ] 
no1000g2$ID[which(is.na(no1000g2[, variant]))]

no1000g2$ID[which(is.na(no1000g2[, variant]))] %>% length()

# variant name
final.scoring.file %>% filter(var.id %in% names(which(colSums(is.na(final.dos))>0)))


### var 2
variant <- names(which(colSums(is.na(final.dos))>0))[2]
summary(final.dos[, variant]) 

# print IDs with missing dosage
final.dos$ID[which(is.na(final.dos[, variant]))]
final.dos$ID[which(is.na(final.dos[, variant]))] %>% length()

# do all missing dosages have HG IDs (1000G)? (no)
length(grep("HG", 
           final.dos$ID[which(is.na(final.dos[, variant]))]))

# exclude 1000G (HG ids) and then print IDs with missing dosage
no1000g <- final.dos[!str_detect(final.dos$ID, "HG"), ] 
no1000g$ID[which(is.na(no1000g[, variant]))]

# exclude 1000G (HG ids) and "NA" ids and then print IDs with missing dosage
no1000g2 <- no1000g[!str_detect(no1000g$ID, "NA"), ] 
no1000g2$ID[which(is.na(no1000g2[, variant]))]

no1000g2$ID[which(is.na(no1000g2[, variant]))] %>% length()

# variant name
final.scoring.file %>% filter(var.id %in% names(which(colSums(is.na(final.dos))>0)))

```

```{bash, eval=FALSE}
grep 57840921 /home/mok36/imputation_code/freeze9b/discovery/9_Minimac4_imputation/discovery-9b-hg38-chr19-imputed.info

bcftools view -r chr19:57840921 /home/shared_data/samoa/wgs/freeze.9b/sites/freeze.9b.chr19.pass_and_fail.sites.bcf -H

grep 57840921 /home/mok36/imputation_code/freeze9b/discovery/3_Affy_discovery_samoa_plink/discovery-samoa-hg38-chr19.bim

plink --bfile /home/mok36/imputation_code/freeze9b/discovery/3_Affy_discovery_samoa_plink/discovery-samoa-hg38-chr19 --freq --snp rs2562052 --out ~/lipids/prs/variant-checking/chr19_57840921
```

Variant with missing genotypes is one that was *not* imputed and was added back from the Affy data. 

The genotyping rate is 0.983328 which translates to exactly 52 missing genotypes.


### rescore variants with NA dosages

```{r}

```


### freeze 8 NA checks

#### check chr 3 (freeze 8 only)

assigning scores for chromosome 3
Summary of PRS on chr 3
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
 0.9737  1.2312  1.2853  1.2825  1.3368  1.5546    1195 


```{r, eval=FALSE}
chr.i <- 3

  load(paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"), verbose=TRUE)
  load(paste0("dosages/final.dos.chr", chr.i))
  
  samoa.scores <- tibble(ID = final.dos$ID,
                      score = 0)
  
  # For each person, extract the vector of dosages and multiple by the vector of effect weights
  weights <- final.scoring.file %>% dplyr::pull(effect_weight)
  
  for(person.i in 1:nrow(samoa.scores)){
    samoa.scores$score[person.i] <- sum(weights * final.dos[person.i, -1 ]) 
  }
  
  summary(samoa.scores$score)

samoa.scores %>% filter(is.na(score)) %>% head()

summary(weights)

# check for missing dosages
#apply(final.dos[, -1], 2, summary)
names(which(colSums(is.na(final.dos))>0))
summary(final.dos[, names(which(colSums(is.na(final.dos))>0))]) # these are genotyped variants that were not imputed

final.scoring.file %>% filter(var.id == names(which(colSums(is.na(final.dos))>0)))
table(final.dos[, names(which(colSums(is.na(final.dos))>0))], useNA = "always") # these are genotyped variants that were not imputed
```
#### check chr 19  (freeze 8 only)

assigning scores for chromosome 19
Summary of PRS on chr 19
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
  8.782  11.304  11.590  11.538  11.836  12.593    1195 


```{r, eval=FALSE}
chr.i <- 19

  load(paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"))
  load(paste0("dosages/final.dos.chr", chr.i))
  
  samoa.scores <- tibble(ID = final.dos$ID,
                      score = 0)
  
  # For each person, extract the vector of dosages and multiple by the vector of effect weights
  weights <- final.scoring.file %>% dplyr::pull(effect_weight)
  
  for(person.i in 1:nrow(samoa.scores)){
    samoa.scores$score[person.i] <- sum(weights * final.dos[person.i, -1 ]) 
  }
  
  summary(samoa.scores$score)

samoa.scores %>% filter(is.na(score)) %>% head()

summary(weights)

# check for missing dosages
#apply(final.dos[, -1], 2, summary)
names(which(colSums(is.na(final.dos))>0))
summary(final.dos[, names(which(colSums(is.na(final.dos))>0))]) # these are genotyped variants that were not imputed

final.scoring.file %>% filter(var.id == names(which(colSums(is.na(final.dos))>0)))
table(final.dos[, names(which(colSums(is.na(final.dos))>0))], useNA = "always") # these are genotyped variants that were not imputed

```

#### resolve missing genotypes (freeze 8 only)

What to do with these two variants? 
Drop from PRS or score them only on the people who are genotyped?


For now, drop from the PRS.

##### rescore chr 3 (freeze 8 only)
```{r, eval=FALSE}
chr.i <- 3

  load(paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"), verbose=TRUE)
  load(paste0("dosages/final.dos.chr", chr.i))
  
  samoa.scores <- tibble(ID = final.dos$ID,
                      score = 0)
  
  exclude <- names(which(colSums(is.na(final.dos))>0))

  final.dos <- final.dos[, which(colSums(is.na(final.dos))==0)] # only keep variants without missing data

  # For each person, extract the vector of dosages and multiple by the vector of effect weights
  weights <- final.scoring.file %>% filter(var.id != exclude) %>% dplyr::pull(effect_weight)
  
  for(person.i in 1:nrow(samoa.scores)){
    samoa.scores$score[person.i] <- sum(weights * final.dos[person.i, -1 ]) 
  }
  
  summary(samoa.scores$score)

  save(samoa.scores, file=paste0("prs-samoan-scores/samoa.prs.chr", chr.i, ".RData"))

```

##### rescore chr 19 (freeze 8 only)
```{r, eval=FALSE}
chr.i <- 19

  load(paste0("scoring-files/final.scoring.file.chr", chr.i, ".RData"), verbose=TRUE)
  load(paste0("dosages/final.dos.chr", chr.i))
  
  samoa.scores <- tibble(ID = final.dos$ID,
                      score = 0)
  
  exclude <- names(which(colSums(is.na(final.dos))>0))

  final.dos <- final.dos[, which(colSums(is.na(final.dos))==0)] # only keep variants without missing data

  # For each person, extract the vector of dosages and multiple by the vector of effect weights
  weights <- final.scoring.file %>% filter(var.id != exclude) %>% dplyr::pull(effect_weight)
  
  for(person.i in 1:nrow(samoa.scores)){
    samoa.scores$score[person.i] <- sum(weights * final.dos[person.i, -1 ]) 
  }
  
  summary(samoa.scores$score)

  save(samoa.scores, file=paste0("prs-samoan-scores/samoa.prs.chr", chr.i, ".RData"))

```

## What percent of PRS variants have MAF = 0 in Samoans?

```{r}
maf.counts <- data.frame(chr = 1:22,
                         n.snp = rep(NA, 22),
                         n.maf0 = rep(0, 22))
for(i in 1:22){
  load(paste0("scoring-files/final.scoring.file.chr", i, ".RData"), verbose=FALSE)
  maf.counts$n.snp[i] <- nrow(final.scoring.file)
  maf.counts$n.maf0[i] <- sum(final.scoring.file$maf == 0, na.rm = T) # 2 vars on chr 19 have missing MAF info
}

sum(maf.counts$n.snp)
sum(maf.counts$n.maf0)
sum(maf.counts$n.maf0)/sum(maf.counts$n.snp)
```


# PRS evaluation in Samoans

## Merge in all chromosomes together
```{r}
load("prs-samoan-scores/samoa.prs.chr1.RData")
names(samoa.scores)[2] <- "score.chr1"

samoa.scores.all <- samoa.scores 


for(chr.i in c(2:22)){
  load(paste0("prs-samoan-scores/samoa.prs.chr", chr.i, ".RData"))
  names(samoa.scores)[2] <- paste0("score.chr", chr.i)
  samoa.scores.all <- left_join(samoa.scores.all, samoa.scores, by = "ID")
}

dim(samoa.scores.all)

samoa.scores.all$prs <- rowSums(samoa.scores.all[,-1])
summary(samoa.scores.all$prs)
sd(samoa.scores.all$prs)
```



## Distribution of PRS in Samoans
```{r}
ggplot(data=samoa.scores.all, aes(x=prs)) + 
  geom_density() + 
  theme_bw()
```


## Association with LDL

Merge with phenotype data

```{bash, eval=FALSE}
#ln -s /home/jnc35/gwas/imputed/imputed_freeze8/LDL_C2-model1/LDL_C2-phenotype.RData

ln -s /home/jnc35/gwas/imputed/imputed_freeze9/discovery/Cholesterol2-model1/Cholesterol2-phenotype.RData

ln -s /home/jnc35/gwas/imputed/imputed_freeze9/discovery/population_structure/data/samoa-hg38-pcair-round2_pcair.RData



```


```{r}
load("Cholesterol2-phenotype.RData", verbose=TRUE)

head(pData(annot))

pca <- get(load("./samoa-hg38_pcair-round2_pcair.RData"))
pcs <- as.data.frame(pca$vectors)
pcs$sample.id <- row.names(pcs)

tmp.df <- left_join(samoa.scores.all, pData(annot), by = c("ID" = "sample.id"))
samoa.prs.df <- left_join(tmp.df, pcs[,c("sample.id", "V1", "V2", "V3")], by = c("ID" = "sample.id"))

samoa.prs.df
```

### Unadjusted model
```{r}
ggplot(data = samoa.prs.df, aes(y=Cholesterol2, x=prs)) + 
  geom_point(alpha=0.3, color="#004D85") + 
  geom_smooth(color="gray20", se=FALSE) + 
  theme_classic() +
  xlab("Polygenic Risk Score") + 
  ylab("Total Cholesterol (mg/dL)") +
  scale_x_continuous(breaks=seq(43,50)) +
  theme(text=element_text (size=32))

ggsave("PRS-Cholesterol2_scatterplot.png", width=16, height=10, units="in")


m1 <- lm(Cholesterol2 ~ prs, data = samoa.prs.df)
summary(m1)

```

didn't end up uding this. ignore stuff with norm 

```{r}
samoa.prs.df$PRS_norm <- RankNorm(samoa.prs.df$prs)
m1.n <- lm(Cholesterol2 ~ PRS_norm, data = samoa.prs.df)
summary(m1.n)
```


### Adjust for age, sex, and PCAs

```{r}
m2 <- lm(Cholesterol2 ~ prs + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = samoa.prs.df)
summary(m2)
confint(m2)

# p-value for prs
beta <- summary(m2)$coefficients["prs", "Estimate"]
beta
confint(m2)
tstat <- summary(m2)$coefficients["prs", "t value"]
2*pt(abs(tstat), df=2849, lower.tail=FALSE)

#m2.n <- lm(Cholesterol2 ~ PRS_norm + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = samoa.prs.df)
#summary(m2.n)
```


### R2 - PRS evaluation metric

```{r}
rsq.partial(objF = m2)

rsq.partial(objF = m2)$partial.rsq[1]

#rsq.partial(objF = m2.n)

#rsq.partial(objF = m2.n)$partial.rsq[1]
```
**The Partial R2 for the PRS in Samoans in 10.2%. ** 

Evaluation of PRS in Graham et al 2021:

"For each individual in the testing cohorts, PRS were calculated as the sum of the dosages multiplied by the given weight at each variant."  

"Model fit was assessed using the adjusted R2 of a linear model for LDL-C value at initial assessment adjusted for cholesterol medication (divided by 0.7 to estimate pre-medication levels) with sex, batch, age at initial assessment and PCs 1–4 as covariates (Supplementary Tables 21–23)."  

"Supplementary Table 23: Prediction of LDL-C in MGI individuals based on varying numbers of PCs included in the model
The polygenic score was normalized within each ancestry group separately"  

"Adjusted R2 values were reported for each cohort and ancestry group, with 95% confidence intervals for the adjusted R2 values calculated using bootstrapping."  

"The PRS developed from the multi-ancestry meta-analysis consistently showed the best or near-best performance in each group tested, with improved or comparable predictions relative to ancestry-matched scores (adjusted R2=0.10–0.16; Fig. 3, Supplementary Table 17, Extended Data Fig. 7)."  


#### Bootstrap CI for R2

```{r}
set.seed(06082022)
n.boot <- 1000
boot.r2 <- rep(NA, times=n.boot)

for(i in 1:n.boot){
  boot.df <- samoa.prs.df[sample(1:nrow(samoa.prs.df), nrow(samoa.prs.df), replace=TRUE),]
  boot.m2 <- lm(Cholesterol2 ~ prs + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = boot.df)
  boot.r2[i] <- rsq.partial(objF = boot.m2)$partial.rsq[1]
}

hist(boot.r2)
quantile(boot.r2, c(0.025, 0.975)) # 95% bootstrap CI for R2



```

### Plot for Poster


Recreate Fig 3 from Graham et al 2021 put add in Samoan data

https://www.nature.com/articles/s41586-021-04064-3/figures/3

```{r}
graham <- read_xlsx("41586_2021_4064_MOESM4_ESM.xlsx", sheet = "ST.17 PRS Validation", skip = 4)
graham

# extract R2 from the all-ancestry PRS evaluation
graham <- graham %>% filter(Risk_score == "ALL") 

# does it match the groups in Fig 3? 
graham %>% pull(ancestry) %>% table()

# almost - there is 1 more AFR
graham

# In this table AADM cohort has an overall and east/west split and AWI-Gen has an overall and East/South/West split
graham %>% pull(cohort) %>% table()

# drop the combined points for AADM and AWI-GEN (not in the Fig 3 from the Graham paper)
graham <- graham %>% filter( !(ancestry == "AFR" & cohort == "AADM") )
graham <- graham %>% filter( !(ancestry == "AFR" & cohort == "AWI-Gen") )
graham
graham %>% pull(ancestry) %>% table() # matches Fig 3

newfig <- graham %>% add_row(Risk_score = NA,
                             `LDL value type` = NA,
                             adj_R2 = 0.1017,
                             Lower_95_CI = 0.0823,
                             Upper_95_CI = 0.1242,
                             ancestry = "Samoan", 
                             cohort = "OLaGA")

newfig %>% select(ancestry, cohort)

newfig <- newfig %>% mutate(label = case_when(ancestry == "AFR (East)" & cohort == "AADM" ~ "AADM\nKenya",
                                              ancestry == "AFR (West)" & cohort == "AADM" ~ "AADM\nNigeria,\nGhana",
                                              ancestry == "AFR (East)" & cohort == "AWI-Gen-East" ~ "AWI-Gen\nKenya",
                                              ancestry == "AFR (South)" & cohort == "AWI-Gen-South" ~ "AWI-Gen\nS.Africa",
                                              ancestry == "AFR (West)" & cohort == "AWI-Gen-West" ~ "AWI-Gen\nBurkina Faso,\nGhana",
                                              ancestry == "SAS" & cohort == "ELGH" ~ "ELGHa",
                                              ancestry == "EAS" & cohort == "KoGES" ~ "KoGES",
                                              ancestry == "AFRAMR" & cohort == "MGI" ~ "MGI\nAFRAMR",
                                              ancestry == "EUR" & cohort == "MGI" ~ "MGI\nEUR",
                                              ancestry == "AFRAMR" & cohort == "MVP" ~ "MVP\nAFRAMR",
                                              ancestry == "ASN" & cohort == "MVP" ~ "MVP\nASN",
                                              ancestry == "EUR" & cohort == "MVP" ~ "MVP\nEUR",
                                              ancestry == "HIS" & cohort == "MVP" ~ "MVP\nHIS",
                                              ancestry == "AFRAMR" & cohort == "PMBB" ~ "PMBB",
                                              ancestry == "EAS" & cohort == "ToMMo" ~ "ToMMo",
                                              ancestry == "Samoan" & cohort == "OLaGA" ~ "Samoa"))


newfig <- newfig %>% mutate(ancestry2 = factor(str_replace_all(ancestry, c("AFRAMR" = "African American",
                                                                    "AFR" = "African",
                                                                    "SAS" = "South Asian", 
                                                                    "EAS" = "East Asian", 
                                                                    "ASN" = "Asian", 
                                                                    "EUR" = "European",
                                                                    "HIS" = "Hispanic")),
                                               levels = c("African (East)", "African (West)", "African (South)", "African American", 
                                                          "European", "Hispanic", "Asian", "East Asian", "South Asian", "Samoan")))

 newfig <- newfig %>% mutate(label2 = factor(label, levels = newfig$label[order(newfig$ancestry2)]))


ggplot(newfig, aes(x=label2, y=adj_R2, color=ancestry2)) + 
  geom_errorbar(aes(ymin=Lower_95_CI, ymax=Upper_95_CI)) +
  geom_point(size=2) +
  scale_y_continuous(labels = c(0,0.05,0.10,0.15,0.20), limits = c(0, 0.20)) +
  ylab(expression(R^2)) + 
  xlab("") + 
  theme_bw() + 
  scale_color_discrete(name = "Ancestry") +
  theme(text=element_text (size=16))

#+
#  scale_color_brewer(palette = "Paired")
#manual(breaks = c("African (East)", "African (West)", "African (South)", "African American", 
#                                "Asian", "East Asian", "South Asian",
#                                "European", "Hispanic", "Samoan"),
#                     values = c(PuRd[]"#c51b7d", "#e41a1c", "#e41a1c", "#e41a1c",
#                                "#b8e186", "#7fbc41", "#4d9221", 
#                                "#3288bd", "#984ea3", "#ff7f00"))


#ggsave("PRS-R2-comparison.png", width=16, height=10, units="in")


newfig <- newfig %>% mutate(label3 = case_when(ancestry == "AFR (East)" & cohort == "AADM" ~ "AADM (Kenyan)",
                                              ancestry == "AFR (West)" & cohort == "AADM" ~ "AADM (Nigerian, Ghanaian)",
                                              ancestry == "AFR (East)" & cohort == "AWI-Gen-East" ~ "AWI-Gen (Kenyan)",
                                              ancestry == "AFR (South)" & cohort == "AWI-Gen-South" ~ "AWI-Gen (South African)",
                                              ancestry == "AFR (West)" & cohort == "AWI-Gen-West" ~ "AWI-Gen (Burkinabe, Ghanaian)",
                                              ancestry == "SAS" & cohort == "ELGH" ~ "ELGH (British Pakistani and Bangladeshi)",
                                              ancestry == "EAS" & cohort == "KoGES" ~ "KoGES (Korean)",
                                              ancestry == "AFRAMR" & cohort == "MGI" ~ "MGI (African American)",
                                              ancestry == "EUR" & cohort == "MGI" ~ "MGI (European American)",
                                              ancestry == "AFRAMR" & cohort == "MVP" ~ "MVP (African American)",
                                              ancestry == "ASN" & cohort == "MVP" ~ "MVP (Asian American)",
                                              ancestry == "EUR" & cohort == "MVP" ~ "MVP  (European American)",
                                              ancestry == "HIS" & cohort == "MVP" ~ "MVP (Hispanic American)",
                                              ancestry == "AFRAMR" & cohort == "PMBB" ~ "PMBB (African American)",
                                              ancestry == "EAS" & cohort == "ToMMo" ~ "ToMMo (Japanese)",
                                              ancestry == "Samoan" & cohort == "OLaGA" ~ "OLaGA (Samoan)"))

newfig <- newfig %>% mutate(label4 = factor(label3, levels = newfig$label3[order(newfig$ancestry2)]))

ggplot(newfig, aes(x=label4, y=adj_R2, color=ancestry2)) + 
  geom_errorbar(aes(ymin=Lower_95_CI, ymax=Upper_95_CI), size=2) +
  geom_point(size=4) +
  scale_y_continuous(labels = c(0,0.05,0.10,0.15,0.20), limits = c(0, 0.20)) +
  ylab(expression(R^2)) + 
  xlab("") + 
  theme_bw() + 
  scale_color_discrete(name = "Ancestry") +
  coord_flip() + 
  theme(text=element_text (size=24),
        legend.position = "none")

ggsave("PRS-R2-comparison_flip.png", width=16, height=10, units="in")



```


# Session Information

```{r}
sessionInfo()
```
