---
title: "Graham et al 2021 LDL-C PRS Performance in Samoans"
author: "Toni-Ann Yapp and Jenna C. Carlson"
date: "`r format(Sys.time(), '%B %d, %Y, %R')`"
output: 
 html_document:
    toc: true
    number_sections: true
    code_folding: show
editor_options: 
  chunk_output_type: inline
---


```{r,echo=FALSE,message=FALSE,warning=FALSE}
require(knitr)
# Set so that long lines in R will be wrapped:
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```


Goal: assess the performance of the multi-ancestry polygenic risk score for LDL-C constructed in Graham et al 2021 (https://www.nature.com/articles/s41586-021-04064-3) in the Samoan discovery cohort.

PRS weights from: https://www.pgscatalog.org/score/PGS000889/

This PRS was developed using GWAS from ALL ancestries in the analyses (1,088,526 individuals of the following ancestries: African unspecified, East Asian, European, Hispanic or Latin American, South Asian).

It was developed using pruning and thresholding of GWAS results (r2=0.1, distance=500kb, p=5e-4) and contains 9,009 variants in build hg19, weighted by effect estimates (betas).


# Load Libraries

```{r load_libraries,message=FALSE}
library(tidyverse)
library(SeqArray)
library(SeqVarTools)
library(TopmedPipeline)
library(Biobase) # annotated data frame
library(rsq)
library(RNOmni)
library(readxl)

```

# Input directory and files 

```{r}
# Print the working directory
getwd()
```
Symbolic links

```{bash, eval=FALSE}
# Example symlinks for local environment (edit with your own paths)
# ln -s /path/to/freeze9b/replication/14_replication_make_gds gds
# ln -s /path/to/replication_gwas_hg38_PC-AiR_round2_pcair.RData
# ln -s /path/to/LDLCT3-model1/LDLCT3-phenotype.RData
# ln -s /path/to/phenotypes-analysis-discovery.RData

```

#CREATE PGS IN MASTER FILE

# Extract variants from Freeze9 GDS files and create scoring files IN SCRIPTS

# Assign Individual Scores IN SCRIPTS

# What percent of PRS variants have MAF = 0 in Samoans?

```{r}
maf.counts <- data.frame(chr = 1:22,
                         n.snp = rep(NA, 22),
                         n.maf0 = rep(0, 22))
for(i in 1:22){
  load(paste0("~/PRS/LDL_1990/scoring-files/final.scoring.file.chr", i, ".RData"), verbose=FALSE)
  maf.counts$n.snp[i] <- nrow(final.scoring.file)
  maf.counts$n.maf0[i] <- sum(final.scoring.file$maf == 0, na.rm = T) # 2 vars on chr 19 have missing MAF info
}

sum(maf.counts$n.snp)
sum(maf.counts$n.maf0)
sum(maf.counts$n.maf0)/sum(maf.counts$n.snp)
```


# PRS evaluation in Samoans

## Merge in all chromosomes together
```{r}
load("~/PRS/LDL_1990/prs-samoan-scores/samoa.prs.chr1.RData")
names(samoa.scores)[2] <- "score.chr1"

samoa.scores.all <- samoa.scores 


for(chr.i in c(2:22)){
  load(paste0("~/PRS/LDL_1990/prs-samoan-scores/samoa.prs.chr", chr.i, ".RData"))
  names(samoa.scores)[2] <- paste0("score.chr", chr.i)
  samoa.scores.all <- left_join(samoa.scores.all, samoa.scores, by = "ID")
}

dim(samoa.scores.all)

samoa.scores.all$prs <- rowSums(samoa.scores.all[,-1])
summary(samoa.scores.all$prs)
sd(samoa.scores.all$prs)
```


## Association with LDL

Merge with phenotype data


```{r}
# Update path to your local phenotype file
load("path/to/LDLCT3-phenotype.RData", verbose=TRUE)

head(pData(annot))

# Update path to your local PC-AiR file
pca <- get(load("path/to/replication_gwas_hg38_PC-AiR_round2_pcair.RData"))
pcs <- as.data.frame(pca$vectors)
pcs$sample.id <- row.names(pcs)

tmp.df <- left_join(samoa.scores.all, pData(annot), by = c("ID" = "sample.id"))
samoa.prs.df <- left_join(tmp.df, pcs[,c("sample.id", "V1", "V2", "V3")], by = c("ID" = "sample.id"))

samoa.prs.df= samoa.prs.df[samoa.prs.df$polity == 1, ]

ldl.obs <- complete.cases(samoa.prs.df$age, samoa.prs.df$sex, samoa.prs.df$LDLCT3)

m.ldl <- lm(LDLCT3 ~ age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = samoa.prs.df[ldl.obs,], na.action = na.exclude) 

samoa.prs.df[ldl.obs, "ldl.res"] <- RankNorm(residuals(m.ldl))

hist(samoa.prs.df$ldl.res)

ggplot(data=samoa.scores.all, aes(x=prs)) + 
  geom_density() + 
  theme_bw()

write.table(samoa.prs.df, 
            file = "~/PRS/individual_prs/ldl_1990_WS_prs.txt", 
            row.names = FALSE,
            quote = FALSE,
            sep = "\t")



```

### Unadjusted model
```{r}
ggplot(data = samoa.prs.df, aes(y=LDLCT3, x=prs)) + 
  geom_point(alpha=0.3, color="#004D85") + 
  geom_smooth(color="gray20", se=FALSE) + 
  theme_classic() +
  xlab("Polygenic Risk Score") + 
  ylab("LDLCT3 (mg/dL)") +
  scale_x_continuous(breaks=seq(43,50)) +
  theme(text=element_text (size=32))

ggsave("PRS-LDL_1990_scatterplot_WS.png", width=16, height=10, units="in")


m1 <- lm(LDLCT3 ~ prs, data = samoa.prs.df)
summary(m1)

```



### Adjust for age, sex, and PCAs

```{r}
m2 <- lm(LDLCT3 ~ prs + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = samoa.prs.df)
summary(m2)
confint(m2)

# p-value for prs
beta <- summary(m2)$coefficients["prs", "Estimate"]
beta
confint(m2)
tstat <- summary(m2)$coefficients["prs", "t value"]
2*pt(abs(tstat), df=2849, lower.tail=FALSE)

```


### R2 - PRS evaluation metric

```{r}
rsq.partial(objF = m2)

rsq.partial(objF = m2)$partial.rsq[1]

#rsq.partial(objF = m2.n)

#rsq.partial(objF = m2.n)$partial.rsq[1]
```


#### Bootstrap CI for R2

```{r}
set.seed(06082022)
n.boot <- 1000
boot.r2 <- rep(NA, times=n.boot)

for(i in 1:n.boot){
  boot.df <- samoa.prs.df[sample(1:nrow(samoa.prs.df), nrow(samoa.prs.df), replace=TRUE),]
  boot.m2 <- lm(LDLCT3 ~ prs + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, data = boot.df)
  boot.r2[i] <- rsq.partial(objF = boot.m2)$partial.rsq[1]
}

hist(boot.r2)
quantile(boot.r2, c(0.025, 0.975)) # 95% bootstrap CI for R2



```


### Incremental R² - Additional variance explained by PRS

```{r}
# Model with covariates only (no PRS)
m2.base <- lm(LDLCT3 ~ age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, 
              data = samoa.prs.df)

# Model with PRS and covariates (already created above as m2)
# m2 <- lm(LDLCT3 ~ prs + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, 
#          data = samoa.prs.df)

# Calculate R² for both models
r2.base <- summary(m2.base)$r.squared
r2.full <- summary(m2)$r.squared

# Incremental R² (additional variance explained by PRS)
incremental.r2 <- r2.full - r2.base

cat("R² (covariates only):", round(r2.base, 4), "\n")
cat("R² (PRS + covariates):", round(r2.full, 4), "\n")
cat("Incremental R² (PRS contribution):", round(incremental.r2, 4), "\n")
cat("Percentage of additional variance explained by PRS:", 
    round(incremental.r2 * 100, 2), "%\n")

# Likelihood ratio test to assess significance of PRS contribution
lrt <- anova(m2.base, m2, test = "F")
print(lrt)

```

#### Bootstrap CI for Incremental R²

```{r}
set.seed(06082022)
n.boot <- 1000
boot.incremental.r2 <- rep(NA, times = n.boot)

for(i in 1:n.boot){
  boot.df <- samoa.prs.df[sample(1:nrow(samoa.prs.df), nrow(samoa.prs.df), replace = TRUE),]
  
  # Base model (covariates only)
  boot.m2.base <- lm(LDLCT3 ~ age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, 
                     data = boot.df)
  
  # Full model (PRS + covariates)
  boot.m2.full <- lm(LDLCT3 ~ prs + age + age2 + sex + sex_age + sex_age2 + V1 + V2 + V3, 
                     data = boot.df)
  
  # Calculate incremental R²
  boot.incremental.r2[i] <- summary(boot.m2.full)$r.squared - summary(boot.m2.base)$r.squared
}

hist(boot.incremental.r2, 
     main = "Bootstrap Distribution of Incremental R²",
     xlab = "Incremental R²",
     col = "#004D85")

# 95% bootstrap CI for incremental R²
boot.ci <- quantile(boot.incremental.r2, c(0.025, 0.975))
cat("\n95% Bootstrap CI for Incremental R²:\n")
cat("Lower bound:", round(boot.ci[1], 4), "\n")
cat("Upper bound:", round(boot.ci[2], 4), "\n")

```

### Save R2 values for Excel updates

```{r}
# Extract Partial R2 and CI
partial_r2_value <- rsq.partial(objF = m2)$partial.rsq[1]
partial_r2_ci <- quantile(boot.r2, c(0.025, 0.975))

# Extract Incremental R2 and CI
incremental_r2_value <- incremental.r2
incremental_r2_ci <- boot.ci

# Save to CSV for Excel update script
r2_results <- data.frame(
  Cohort = "1990",
  Trait = "LDL-C",
  Polity = "WS",
  Year = 1995,
  Partial_R2 = partial_r2_value,
  Partial_R2_LCI = partial_r2_ci[1],
  Partial_R2_UCI = partial_r2_ci[2],
  Incremental_R2 = incremental_r2_value,
  Incremental_R2_LCI = incremental_r2_ci[1],
  Incremental_R2_UCI = incremental_r2_ci[2],
  stringsAsFactors = FALSE
)

write.csv(r2_results, file = "../LDL_r2_results_1990_WS.csv", row.names = FALSE)
cat("\nR2 values saved to: LDL_r2_results_1990_WS.csv\n")
print(r2_results)
```


# Session Information

```{r}
sessionInfo()
```
